Inference using context-dependent models of mutation

### 2017-06-05 writeups/context-dependence.tex
writing

\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\calS}{\mathcal{S}}  % set of states
\newcommand{\calT}{\mathcal{T}}  % set of transition triples
\newcommand{\calP}{\mathcal{P}}  % set of selection pairs
\newcommand{\nA}{\mbox{A}}  % nucleotides:
\newcommand{\nC}{\mbox{C}}
\newcommand{\nG}{\mbox{G}}
\newcommand{\nT}{\mbox{T}}
\newcommand{\nW}{\mbox{W}}
\newcommand{\nR}{\mbox{R}}
\newcommand{\nY}{\mbox{Y}}
\newcommand{\join}{\oplus}  % matches
\newcommand{\st}{\colon}  % such that
\newcommand{\var}{\mathop{\mbox{var}}}
\newcommand{\cov}{\mathop{\mbox{cov}}}
\newcommand{\given}{\;\mid\;}
\newcommand{\like}{\mathcal L}
\newcommand{\loglike}{\ell}
\newcommand{\alike}{\widetilde{\like}}
\newcommand{\aloglike}{\widetilde{\loglike}}

@@

In the context of genomics,
it is well-known that certain short nucleotide sequences
are in many organisms much more, or less, abundant than expected by chance \citep{burge1992underrepresentation},
due to the combined effects of context dependence of the nucleotide mutation process,
selective constraints on the function of the sequence,
and other processes such as biased gene conversion \citep{duret2009biased}.
Indeed, molecular studies have demonstrated that the spectrum of new mutations in humans
is highly context dependent \citep{schaibley2013influence}
and that multinucleotide substitutions are relatively common \citep{schrider2011pervasive,terekhanova2013prevalence,harris2013errorprone}.
In humans, the molecular processes underlying mutations are best studied in contexts of intense somatic mutation:
cancer and the hypervariable regions that generate immune system diversity \citep{evolution_within},
which provide quite different cellular contexts but share commonalities due to shared molecular machinery.
In general, substitution rates depend on sequence context due to the interaction of a great many factors including:
(a) likelihood of DNA damage or errors in cell division (e.g., UV damage \citep{UV} or non-homologous end joining \citep{NHEJ});
(b) activity levels and structural properties of the enzymes responsible for repair (e.g., base excision repar \citep{BER} and Y-family polymerases \citep{polymerase_review});
and (c) any functional effects of an error (e.g., lethal mutations will not be seen in a population census).
(We are concerned with effects that are \emph{homogeneous} across the sequence,
so do not consider further the many constraints on protein-coding sequence \citep[reviewed in][]{thorne2007protein}.)
Some of these effects -- particularly, activity levels of different enzymes in the germline --
will likely vary over evolutionary time,
and so comparisons between widely separated species will identify time-averaged mutation rates.
Recent studies have used nonnegative matrix factorization to identify sets of mutational patterns
co-occurring in comparisons (a) between different cancer types \citep{cancer_scan} and (b) between different human populations \citep{harris,matthieson},
to disentangle distinct signals presumably coming from the action of distinct sources of mutation.

@@

      \item[(repair)] In addition to the above, a collection of motifs associated with known biochemical causes
          curated from the literature:
          deamination due to repair of UV-induced pyramidine dimers ($\mu_{UV}$: (\nT/\nC)\nC $\to$ (\nT/\nC)\nT and \nC\nC$\to$\nT\nT, \citet{uv_damage});
          an interaction between CpG and pyramidine dimers ($\mu_{mUV}$: \nT\nC\nG$\to$\nT\nT\nG and \nC\nC\nG$\to$\nC\nT\nG, \citet{cpg_review});
          repair of guanine adducts at CpG sites ($\mu_g$: \nC\nG$\to$\nC\nT, \citet{tobacco});
          deamination by AID ($\mu_{AID}$: \nW\nR\nC\nY$\to$\nW\nR\nT\nY, \nW is \nA or \nT; \nR is \nA or \nG; \nY is \nT or \nC, \citet{AID});
          deamination by members of the APOBEC family ($\mu_{\APOBEC}$: \nT\nC\nW$\to$\nT\nT\nW or \nT\nC\nW$\to$\nT\nG\nW, \citet{APOBEC});
          and repair by error-prone polymerases pol $\eta$ ($\mu_\eta$: \nT\nA$\to$\nT\nG, \citet{pol_eta});
          and pol $\iota$ ($\mu_{\iota}$: \nA\nA$\to$\nG\nT and \nA\nA$\to$\nA\nG \citet{pol_iota}):

          \noindent
      \begin{center}
          \begin{tabular}{c@{\;\;$\to$\;\;}c@{\;\; {\small at rate}\;\; }c|c@{\;\;$\to$\;\;}c@{\;\; {\small at rate}\;\; }c|c@{\;\;$\to$\;\;}c@{\;\; {\small at rate}\;\; }c}
            \nT\nC     &   \nT\nT      &  $\mu_{UV1}$    & \nC\nG       &   \nC\nT       &  $\mu_{g}$       & \nA\nA\nC\nT  &   \nA\nA\nT\nT   &  $\mu_{\AID}$ \\
            \nG\nA     &   \nA\nA      &  $\mu_{UV1}$    & \nC\nG       &   \nA\nG       &  $\mu_{g}$       & \nA\nA\nC\nC  &   \nA\nA\nT\nC   &  $\mu_{\AID}$ \\
            \nC\nC     &   \nC\nT      &  $\mu_{UV2}$    & \nT\nC\nA    &   \nT\nT\nA    &  $\mu_{\APOBEC}$ & \nA\nG\nC\nT  &   \nA\nG\nT\nT   &  $\mu_{\AID}$ \\
            \nG\nG     &   \nA\nG      &  $\mu_{UV2}$    & \nT\nC\nT    &   \nT\nT\nT    &  $\mu_{\APOBEC}$ & \nA\nG\nC\nC  &   \nA\nG\nT\nC   &  $\mu_{\AID}$ \\
            \nC\nC     &   \nT\nT      &  $\mu_{UV3}$    & \nT\nC\nA    &   \nT\nG\nA    &  $\mu_{\APOBEC}$ & \nA\nA\nC\nT  &   \nA\nA\nT\nT   &  $\mu_{\AID}$ \\
            \nG\nG     &   \nA\nA      &  $\mu_{UV3}$    & \nT\nC\nT    &   \nT\nG\nT    &  $\mu_{\APOBEC}$ & \nA\nA\nC\nC  &   \nA\nA\nT\nC   &  $\mu_{\AID}$ \\
            \nT\nC\nG  &   \nT\nT\nG   &  $\mu_{mUV}$    & \nT\nG\nA    &   \nT\nA\nA    &  $\mu_{\APOBEC}$ & \nA\nG\nC\nT  &   \nA\nG\nT\nT   &  $\mu_{\AID}$ \\
            \nC\nC\nG  &   \nC\nT\nG   &  $\mu_{mUV}$    & \nA\nG\nA    &   \nA\nA\nA    &  $\mu_{\APOBEC}$ & \nA\nG\nC\nC  &   \nA\nG\nT\nC   &  $\mu_{\AID}$ \\
            \nC\nG\nA  &   \nC\nA\nA   &  $\mu_{mUV}$    & \nT\nG\nA    &   \nT\nC\nA    &  $\mu_{\APOBEC}$ & \nT\nA\nC\nT  &   \nT\nA\nT\nT   &  $\mu_{\AID}$ \\
            \nC\nG\nG  &   \nC\nA\nG   &  $\mu_{mUV}$    & \nA\nC\nA    &   \nA\nC\nA    &  $\mu_{\APOBEC}$ & \nT\nA\nC\nC  &   \nT\nA\nT\nC   &  $\mu_{\AID}$ \\
            \nT\nA     &   \nT\nG      &  $\mu_{\eta}$   & \nA\nA       &    \nA\nG      &  $\mu_{\iota1}$  & \nT\nG\nC\nT  &   \nT\nG\nT\nT   &  $\mu_{\AID}$ \\
            \nT\nA     &   \nC\nA      &  $\mu_{\eta}$   & \nT\nT       &    \nC\nT      &  $\mu_{\iota1}$  & \nT\nG\nC\nC  &   \nT\nG\nT\nC   &  $\mu_{\AID}$ \\
            \nA\nA     &   \nG\nT      &  $\mu_{\iota2}$ & \nT\nG\nC\nT &  \nT\nG\nT\nT  &  $\mu_{\AID}$    & \nT\nA\nC\nT  &   \nT\nA\nT\nT   &  $\mu_{\AID}$ \\
            \nT\nT     &   \nA\nC      &  $\mu_{\iota2}$ & \nT\nG\nC\nC &  \nT\nG\nT\nC  &  $\mu_{\AID}$    & \nT\nA\nC\nC  &   \nT\nA\nT\nC   &  $\mu_{\AID}$ 
        \end{tabular}
      \end{center}

% [ [ "TC", "TT" ], [ "GA", "AA" ] ], UV1
% [ [ "CC", "CT" ], [ "GG", "AG" ] ], UV2
% [ [ "CC", "TT" ], [ "GG", "AA" ] ], UV3
% [ [ "TCG", "TTG" ], [ "CCG", "CTG" ], [ "CGA", "CAA" ], [ "CGG", "CAG" ] ], mUV
% [ [ "CG", "CT" ], [ "CG", "AG" ] ],
% [ [ "AACT", "AATT" ], [ "AACC", "AATC" ], [ "AGCT", "AGTT" ], [ "AGCC", "AGTC" ], [ "TACT", "TATT" ], [ "TACC", "TATC" ], [ "TGCT", "TGTT" ], [ "TGCC", "TGTC" ],
%   [ "AGTT", "AATT" ], [ "GGTT", "GATT" ], [ "AGCT", "AACT" ], [ "GGCT", "GACT" ], [ "AGTA", "AATA" ], [ "GGTA", "GATA" ], [ "AGCA", "AACA" ], [ "GGCA", "GACA" ] ],
% [ [ "TCA", "TTA" ], [ "TCT", "TTT" ], [ "TCA", "TGA" ], [ "TCT", "TGT" ],
%   [ "TGA", "TAA" ], [ "AGA", "AAA" ], [ "TGA", "TCA" ], [ "AGA", "ACA" ] ],
% [ [ "TA", "TG" ], [ "TA", "CA" ] ],
% [ [ "AA", "GT" ], [ "TT", "AC" ] ],
% [ [ "AA", "AG" ], [ "CT", "TT" ] ]

@@

@@

\textbf{To-do:}

3. Analyze human data with shape model, Bayesian.

The method works well to find point estimates of the parameters.
Obtaining posterior density estimates using the approximation likelihood function on a subset of patterns
works in some cases;
in other cases the likelihood surface is still too peaked, underestimating the uncertainty.
(but counting in nonoverlapping windows fixes the peakiness problem)

@@

\begin{figure}
    \begin{center} 
        \includegraphics[width=\textwidth]{writeup-plots/biochem_mcmc_posteriors}
    \end{center} 
    \caption{
        Estimated average mutation rates
        for the motifs of Table \ref{tab:hominid_model}
        since the common ancestor of humans and chimpanzee.
        Rates are scaled so that the mean time to common ancestor is 0.001 units --
        if we take this to be 6 million years \citep{scally2012insights}
        then rates are in units of mutations per 6,000 years.
        Shown are posterior medians (points) and 95\% credible intervals (lines, which are quite small in all cases).
        Figure \ref{sfig:biochem_results_zoom} shows the smaller rates in more detail.
        \label{fig:biochem_results}}
\end{figure}

@@

>>>>

\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\calS}{\mathcal{S}}  % set of states
\newcommand{\calT}{\mathcal{T}}  % set of transition triples
\newcommand{\calP}{\mathcal{P}}  % set of selection pairs
\newcommand{\nA}{\mbox{A}}  % nucleotides:
\newcommand{\nC}{\mbox{C}}
\newcommand{\nG}{\mbox{G}}
\newcommand{\nT}{\mbox{T}}
\newcommand{\nS}{\mbox{S}} % C or G
\newcommand{\nW}{\mbox{W}} % A or T
\newcommand{\nR}{\mbox{R}} % purine, A or G
\newcommand{\nY}{\mbox{Y}} % pyramidine, C or T
\newcommand{\join}{\oplus}  % matches
\newcommand{\st}{\colon}  % such that
\newcommand{\var}{\mathop{\mbox{var}}}
\newcommand{\cov}{\mathop{\mbox{cov}}}
\newcommand{\given}{\;\mid\;}
\newcommand{\like}{\mathcal L}
\newcommand{\loglike}{\ell}
\newcommand{\alike}{\widetilde{\like}}
\newcommand{\aloglike}{\widetilde{\loglike}}

@@

In the context of genomics,
it is well-known that certain short nucleotide sequences
are in many organisms much more, or less, abundant than expected by chance \citep{burge1992underrepresentation},
due to the combined effects of context dependence of the nucleotide mutation process,
selective constraints on the function of the sequence,
and other processes such as biased gene conversion \citep{duret2009biased,arbeithuber2015crossovers}.
Indeed, molecular studies have demonstrated that the spectrum of new mutations in humans
is highly context dependent \citep{schaibley2013influence}
and that multinucleotide substitutions are relatively common \citep{schrider2011pervasive,terekhanova2013prevalence,harris2013errorprone}.
In humans, the molecular processes underlying mutations are best studied in contexts of intense somatic mutation:
cancer and the hypervariable regions that generate immune system diversity \citep{cobey2015evolution,},
which provide quite different cellular contexts but share commonalities due to shared molecular machinery.
In general, substitution rates depend on sequence context due to the interaction of a great many factors including:
(a) likelihood of DNA damage or errors in cell division (e.g., UV damage \citep{goodman2002errorprone} or crossing over in meiosis \citep{arbeithuber2015crossovers});
(b) activity levels and structural properties of the enzymes responsible for repair (e.g., base excision repar \citep{BER} and Y-family polymerases \citep{polymerase_review});
and (c) any functional effects of an error (e.g., lethal mutations will not be seen in a population census).
(We are concerned with effects that are \emph{homogeneous} across the sequence,
so do not consider further the many constraints on protein-coding sequence \citep[reviewed in][]{thorne2007protein}.)
Some of these effects -- particularly, activity levels of different enzymes in the germline --
will likely vary over evolutionary time,
and so comparisons between widely separated species will identify time-averaged mutation rates.
Recent studies have used nonnegative matrix factorization to identify sets of mutational patterns
co-occurring in comparisons (a) between different cancer types \citep{cancer_scan} and (b) between different human populations \citep{harris,matthieson},
to disentangle distinct signals presumably coming from the action of distinct sources of mutation.

@@

      \item[(repair)] In addition to the above, we included a collection of mutation motifs associated with known DNA lesion/repair pathways
          curated from the literature \citep[reviewed in][]{sale2012yfamily,goodman2013translesion,roberts2014hypermutation,cobey2015evolution,goodman2002errorprone}:
          deamination due to repair of UV-induced pyramidine dimers \citep[$\mu_{UV}$][]{sale2012yfamily,sinha2002uvinduced};
          repair of guanine adducts at CpG sites \citet[$\mu_g$][]{pfeifer2006mutagenesis};
          deamination by AID \citep[$\mu_{AID}$][]{teng2007immunoglobulin,kasar2015wholegenome};
          deamination by members of the APOBEC family \citep[$\mu_\APOBEC$][]{alexandrov2013deciphering};
          and repair by error-prone polymerases pol $\eta$ \citep[$\mu_\eta$][]{alexandrov2013deciphering};
          and pol $\iota$ \citep[$\mu_{\iota}$][]{roberts2014hypermutation,maul2016polymerase}.
          These are denoted here using ambiguity codes 
          \nR{}: \nA{} or \nG{},
          \nY{}: \nC{} or \nT{},
          \nW{}: \nA{} or \nT{}, and
          \nS{}: \nC{} or \nG{};
          so for instance 
          ``\nT\nC\nW$\to$\nT\nT\nW'' 
          means that both 
          \nT\nC\nA$\to$\nT\nT\nA
          and
          \nT\nC\nT$\to$\nT\nT\nT
          share the same mutation rate.
          \noindent
      \begin{center}
          \begin{tabular}{c@{\;\;$\to$\;\;}c@{\;\; {\small at rate}\;\; }c|c@{\;\;$\to$\;\;}c@{\;\; {\small at rate}\;\; }c}
            \nC\nG     &   \nC\nT      &  $\mu_{g}$      & \nA\nA       &   \nA\nG       &  $\mu_{\iota1}$   \\
            \nC\nG     &   \nA\nG      &  $\mu_{g}$      & \nT\nT       &   \nC\nT       &  $\mu_{\iota1}$   \\
            \nT\nC     &   \nT\nT      &  $\mu_{UV}$     & \nA\nA       &   \nG\nT       &  $\mu_{\iota2}$   \\
            \nG\nA     &   \nA\nA      &  $\mu_{UV}$     & \nT\nT       &   \nA\nC       &  $\mu_{\iota2}$   \\
            \nC\nC     &   \nC\nT      &  $\mu_{UV}$     & \nT\nC\nW    &   \nT\nT\nW    &  $\mu_{\APOBEC}$  \\
            \nG\nG     &   \nA\nG      &  $\mu_{UV}$     & \nT\nC\nW    &   \nT\nG\nW    &  $\mu_{\APOBEC}$  \\
            \nC\nC     &   \nT\nT      &  $\mu_{UVd}$    & \nS\nG\nA    &   \nS\nA\nA    &  $\mu_{\APOBEC}$  \\
            \nG\nG     &   \nA\nA      &  $\mu_{UVd}$    & \nS\nG\nA    &   \nS\nC\nA    &  $\mu_{\APOBEC}$  \\
            \nT\nA     &   \nT\nG      &  $\mu_{\eta}$   & \nW\nR\nC\nY &  \nW\nR\nT\nY  &  $\mu_{\AID}$     \\
            \nT\nA     &   \nC\nA      &  $\mu_{\eta}$   & \nR\nG\nY\nS &  \nR\nA\nY\nS  &  $\mu_{\AID}$     
        \end{tabular}
      \end{center}

@@

The mutational motifs included above are far from complete 
-- context-specific affinities and error rates \textit{in vivo} 
are not known for most DNA repair pathways --
but neither do they exhaust the hypotheses available in the literature.
For instance, many enzymes act more than once locally when they act,
such as AID \citep{senavirathne2015activationinduced} or error-prone polymerases involved in repair \citep{maul2016polymerase}.

@@

@@

\begin{figure}
    \begin{center} 
        \includegraphics[width=\textwidth]{writeup-plots/biochem_mcmc_posteriors}
    \end{center} 
    \caption{
        Estimated average mutation rates
        for the motifs of Table \ref{tab:hominid_model}
        since the common ancestor of humans and chimpanzee.
        Rates are scaled so that the mean time to common ancestor is 0.001 units --
        if we take this to be 6 million years \citep{scally2012insights,langergraber2012generation}
        then rates are in units of mutations per 6,000 years.
        Shown are posterior medians (points) and 95\% credible intervals (lines, which are quite small in all cases).
        Figure \ref{sfig:biochem_results_zoom} shows the smaller rates in more detail.
        \label{fig:biochem_results}}
\end{figure}

It may seem tenous to look for mutational signatures associated with cancer in the germline,
or absurd that UV damage could play a significant role.
This is in part due to the ``streetlight effect'' --
we can't look for signatures that are not characterized --
but it is also reasonable that shared repair machinery might induce similar context-dependent errors.
More likely sources of error in the germline include the effects of recombination 
\citep{arbeithuber2015crossovers,myers2010drive}.

@@

\subsection{Longer models}
      \begin{center}
          \begin{tabular}{c@{\;\;$\to$\;\;}c@{\;\; {\small at rate}\;\; }c|c@{\;\;$\to$\;\;}c@{\;\; {\small at rate}\;\; }c|c@{\;\;$\to$\;\;}c@{\;\; {\small at rate}\;\; }c}
            \nT\nC     &   \nT\nT      &  $\mu_{UV1}$    & \nC\nG       &   \nC\nT       &  $\mu_{g}$       & \nA\nA\nC\nT  &   \nA\nA\nT\nT   &  $\mu_{\AID}$ \\
            \nG\nA     &   \nA\nA      &  $\mu_{UV1}$    & \nC\nG       &   \nA\nG       &  $\mu_{g}$       & \nA\nA\nC\nC  &   \nA\nA\nT\nC   &  $\mu_{\AID}$ \\
            \nC\nC     &   \nC\nT      &  $\mu_{UV2}$    & \nT\nC\nA    &   \nT\nT\nA    &  $\mu_{\APOBEC}$ & \nA\nG\nC\nT  &   \nA\nG\nT\nT   &  $\mu_{\AID}$ \\
            \nG\nG     &   \nA\nG      &  $\mu_{UV2}$    & \nT\nC\nT    &   \nT\nT\nT    &  $\mu_{\APOBEC}$ & \nA\nG\nC\nC  &   \nA\nG\nT\nC   &  $\mu_{\AID}$ \\
            \nC\nC     &   \nT\nT      &  $\mu_{UV3}$    & \nT\nC\nA    &   \nT\nG\nA    &  $\mu_{\APOBEC}$ & \nA\nA\nC\nT  &   \nA\nA\nT\nT   &  $\mu_{\AID}$ \\
            \nG\nG     &   \nA\nA      &  $\mu_{UV3}$    & \nT\nC\nT    &   \nT\nG\nT    &  $\mu_{\APOBEC}$ & \nA\nA\nC\nC  &   \nA\nA\nT\nC   &  $\mu_{\AID}$ \\
            \nT\nC\nG  &   \nT\nT\nG   &  $\mu_{mUV}$    & \nT\nG\nA    &   \nT\nA\nA    &  $\mu_{\APOBEC}$ & \nA\nG\nC\nT  &   \nA\nG\nT\nT   &  $\mu_{\AID}$ \\
            \nC\nC\nG  &   \nC\nT\nG   &  $\mu_{mUV}$    & \nA\nG\nA    &   \nA\nA\nA    &  $\mu_{\APOBEC}$ & \nA\nG\nC\nC  &   \nA\nG\nT\nC   &  $\mu_{\AID}$ \\
            \nC\nG\nA  &   \nC\nA\nA   &  $\mu_{mUV}$    & \nT\nG\nA    &   \nT\nC\nA    &  $\mu_{\APOBEC}$ & \nT\nA\nC\nT  &   \nT\nA\nT\nT   &  $\mu_{\AID}$ \\
            \nC\nG\nG  &   \nC\nA\nG   &  $\mu_{mUV}$    & \nA\nC\nA    &   \nA\nC\nA    &  $\mu_{\APOBEC}$ & \nT\nA\nC\nC  &   \nT\nA\nT\nC   &  $\mu_{\AID}$ \\
            \nT\nA     &   \nT\nG      &  $\mu_{\eta}$   & \nA\nA       &    \nA\nG      &  $\mu_{\iota1}$  & \nT\nG\nC\nT  &   \nT\nG\nT\nT   &  $\mu_{\AID}$ \\
            \nT\nA     &   \nC\nA      &  $\mu_{\eta}$   & \nT\nT       &    \nC\nT      &  $\mu_{\iota1}$  & \nT\nG\nC\nC  &   \nT\nG\nT\nC   &  $\mu_{\AID}$ \\
            \nA\nA     &   \nG\nT      &  $\mu_{\iota2}$ & \nT\nG\nC\nT &  \nT\nG\nT\nT  &  $\mu_{\AID}$    & \nT\nA\nC\nT  &   \nT\nA\nT\nT   &  $\mu_{\AID}$ \\
            \nT\nT     &   \nA\nC      &  $\mu_{\iota2}$ & \nT\nG\nC\nC &  \nT\nG\nT\nC  &  $\mu_{\AID}$    & \nT\nA\nC\nC  &   \nT\nA\nT\nC   &  $\mu_{\AID}$ 
        \end{tabular}
      \end{center}
